Type Error Telemetry at Scale

### 2023-09-27 programming23/paper.tex
conclusion: talk alouds, idioms

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Although it appears that creators choose to fix errors,
the fixes may be a side effect of other plans.

@@

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all creators (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    this mode exists.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Creators tend to fix them immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, the \mstrict{} mode clearly needs to relax its data-asset
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights;
    in~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing any source code.
    It was especially useful to know the overlap for old and for current type errors.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time~(\cref{f:indy-session}).
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down~(\cref{f:error-density}).
\end{itemize}

One aspect of the telemetry that we would change for next time is to provide
more metadata.
Luau recently adopted semantic subtyping to reduce false-positive type
errors~\cite{luau-subt}, but the current telemetry has no way to tell if this
internal extension is enabled or not.
Three other aspects are worth rethinking.
First, it would be useful to have fine-grained counts for \FS{} analysis
errors and to know the specific reason behind errors such as \code{TypeMismatch}.
But, this additional data could double the size of telemetry records.
Second, although tracking edit ranges was useful, it made the telemetry system
much more difficult to build and maintain.
Tracking old and current errors at the module level would be more maintainable,
but there is a risk that it is too coarse.
Third, the massive number of syntax errors begs the question of how to reduce
them.
Shifting focus from arbitrary keystrokes to selected ones, such as closing
parentheses or whitespaces, might increase the likelihood of well-formed
code without losing the ``middle of things'' nature of the data.
Another option is to ignore records that have a syntax error in the edit range.

In future work, it would be interesting to build statistical models of
the ``average'' programmer using each typing mode and compare their error
rates.
While one could start modeling with the \mnocheck{} and \mnonstrict{}
data, but there are too few \mstrict{} records at the moment due to our uniform
sampling rate.
More data is needed to experiment with models of \mstrict{}-mode programmers.

\FILL{} future: interviews, accommodate idioms

% \paragraph{Bad: Noisy Data}
% 
% \begin{verbatim}
%  require(foobar)
%   foobar is a module script in the data model
%  require(foo.bar.b)
%   - could be edit in progress
%   - could be renamed module
%   - 
% \end{verbatim}

@@

Thanks to Benjamin Chung for several helpful discussions about data analysis
and effective plotting.
Greenman was supported by
grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.

>>>>

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Although it appears that creators choose to fix errors with their edits,
the fixes may be a side effect of other plans.

@@

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all creators (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Creators tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data-asset
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights.
    In~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing any source code.
    It was especially useful to have overlaps for two sets of type errors: old and current.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time~(\cref{f:indy-session}).
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down~(\cref{f:error-density}).
\end{itemize}

One aspect of the telemetry that we would change for next time is to provide
more metadata.
Luau recently adopted semantic subtyping to reduce false-positive type
errors~\cite{luau-subt}, but the current telemetry has no way to tell if this
internal extension is enabled or not.
Three other aspects are worth rethinking.
First, it would be useful to have fine-grained counts for \FS{} analysis
errors and to know the specific reason behind errors such as \code{TypeMismatch}.
But, this additional data could double the size of telemetry records.
Second, although tracking edit ranges was useful, it made the telemetry system
much more difficult to build and maintain.
Tracking old and current errors at the module level would be easier,
though there is a risk that it is too coarse.
Third, the massive number of syntax errors begs the question of how to reduce
them.
Shifting focus from arbitrary keystrokes to selected ones, such as closing
parentheses or whitespaces, might increase the likelihood of well-formed
code without losing the ``middle of things'' nature of the data.
Another option is to ignore records that have a syntax error in the edit range.

In future work, it would be interesting to build statistical models of
the ``average'' programmer using each typing mode and compare their error
rates.
While one could start modeling with the \mnocheck{} and \mnonstrict{}
data, but there are too few \mstrict{} records at the moment due to our uniform
sampling rate.
The errors that we do see in \mstrict{} mode call for an in-depth analysis via
talk-aloud interviews to discover why type mismatches occur and how how Luau
can better accommodate untyped idioms.

@@

Thanks to Benjamin Chung for several helpful discussions about data analysis
and effective plotting.
Greenman was supported by
NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
